---
title: "Data Wrangling Exercise 01"
author: "Tom Thorpe"
date: "March 27, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Springboard Data Science Foundation Class
Exercise 01 - Data Wrangling - Refine Data

## Objective
Practice using *tidy* and *dplyr* packages to clean up data in a practice dataset.

## Exercise Results
Load the tidy and dplyr libraries for exercise.
```{r}
#library(devtools)
library(tidyr)
library(dplyr)
```

### Identify input and output files
The practice data set was downloaded into a file called, "refine_original.csv" and
the cleaned data will be stored in a file named, "refine_clean.csv".
```{r}
infile = "C:/Users/Tom/git/datasciencefoundation/DataWrangleExer01/refine_original.csv"
outfile = "C:/Users/Tom/git/datasciencefoundation/DataWrangleExer01/refine_clean.csv"  
```

### Read CSV file into a dataframe
```{r}
refineData <- read.csv(file=infile, header=TRUE, sep=",")
```

Lets see what the data looks like:
```{r}
print(refineData)
```

Load into a local data frame and check it's contents.
```{r}
products <- tbl_df(refineData) 
print(products)
```

### Clean Company Name
The first taks to is clean up company names by correcting spellings.
The company names have various mis-spelling that most all start with the first letter of 
the company name. The exception is one entry that starts the spelling of 'philips' with an 'f'.
A regular expression is used to select the appropriate records for each of the company names by 
checking the first character of the company name and change the
name to 'adzo', philips', 'unilever', or 'van houten' 

```{r}
products$company[grepl("^a",products$company,ignore.case=TRUE)] <- "akzo"
products$company[grepl("^p|^f",products$company,ignore.case=TRUE)] <- "philips"
products$company[grepl("^u",products$company,ignore.case=TRUE)] <- "unilever"
products$company[grepl("^v",products$company,ignore.case=TRUE)] <- "van houten"
```
Check the results:
```{r}
print(products,n=30)
```
### Separate Product Code Number
Next separate the product code number into *product_code* and *product_number* columns by using
the *separate* function to split the data at the dash. 
The exercise instructions do not indicate if the *Product.code...number* column is
to be kept or removed, so it is kept by setting the *remove* option to **FALSE**.
```{r}
products <- separate(products,Product.code...number,
                     c("product_code","product_number"),
                     sep="-",
                     remove=FALSE)
```

Check the results:
```{r}
select(products,contains("product")) %>% print(n=25)
```
### Create *product_category* column
A *product_category* column is created to give a descriptive name to the product code.
First the column is created with the mutate function. The value is defaulted to **NA**.
```{r}
products <- mutate(products,product_category=NA)
select(products,contains("product")) %>% print(n=25)
```
Then the column is populated based on the value of the *product_code* using two different logical tests:
```{r}
products$product_category[grepl("p",products$product_code)] <- "Smartphone"
products$product_category[grepl("q",products$product_code)] <- "Tablet"
products$product_category[products$product_code=="v"] <- "TV"
products$product_category[products$product_code=="x"] <- "Laptop"
```
Check the results:
```{r}
select(products,contains("product")) %>% print(n=25)
```

### Create a full address column
To enable Geo-coding, a *full_address* column is added 
by combining the *address*, *city* and *state* columns, with each column separated by a comma.
```{r}
products <- mutate(products,full_address=paste(address,",",city,",",country))
```
Check the results:
```{r}
select(products,c("address","city","country","full_address")) %>% print(n=25)
```
The *full_address* is being truncated. Since the *city* and *country* are the same for each row, lets just look at the *address* and *full_address*.
```{r}
select(products,contains("address")) %>% print(n=25)
```

# Create binary columns for data analysis
New binary category columns are created to allow for easier data analysis. The binary columns have
a value of zero unless the data represented by the column is present for that row.

Four columns are created representing each possible company and the value is defaulted to 0.
```{r}
products <- mutate(products,company_philips=0)
products <- mutate(products,company_akzo=0)
products <- mutate(products,company_van_houten=0)
products <- mutate(products,company_unilever=0)
```

Then the the company category column values are set to 1 if the corresponding company name is present.
```{r}
products$company_akzo[products$company=="akzo"] <- 1
products$company_philips[products$company=="philips"] <- 1
products$company_van_houten[products$company=="van houten"] <- 1
products$company_unilever[products$company=="unilever"] <- 1
```

Check the results.
```{r}
select(products,contains("company")) %>% print(n=25)
```
Next create binary data columns for the product code.
I was wondering if there was a way to create the column and set the value all on the same line and
tried a few ways. I first tried the following and it produced an error:
```{r}
products <- mutate(products,product_tablet2=(if(products$product_category=="Tablet"){1}else{0}))
```
Remove the product_tablet2 column.
```{r}
products <- select(products,-product_tablet2)
```
Next I tried the *ifelse* construct and found it worked and was able to create the column
and set the value in one statement.

```{r}
products <- mutate(products,product_smartphone=ifelse(product_category=="Smartphone",1,0))
products <- mutate(products,product_tv=ifelse(product_category=="TV",1,0))
products <- mutate(products,product_laptop=ifelse(product_category=="Laptop",1,0))
products <- mutate(products,product_tablet=ifelse(product_category=="Tablet",1,0))
```

Check the results using different selects.
```{r}
select(products,contains("product")) %>% print(n=25)
# The below was working
select(products,9,15:18) %>% print(n=25)
# but now getting error:
#   Error in combine_vars(vars, ind_list) : Position must be between 0 and n
# It's working again. I had an error in the mutate functions 
# and was not actually creating the new columns, so the 15:18 range was invalid.
select(products,contains("product"),
       -Product.code...number,
       -product_code,
       -product_number) %>% 
  print(n=25)
select(products,"product_category","product_tablet") %>% print(n=25)
#View(products)
```
Lets look at the final structure for fun.
```{r}
str(products)
```
Save the cleaned data in a CSV file.
```{r}
write.csv(products, file=outfile,row.names=FALSE)
```
That concludes the exercise.